---
title: "Prior elicitation"
author: "Lorenzo Gaborini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prior elicitation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

source('knitr_markup.R')
bmatrix.digits <- 4
```

## Test case

Let's generate a test dataset with known parameters.      
We will use some supporting functions:

```{r}
library(bayessource)

# Generate n random integers in {min, ..., max}
randi <- function(n, min, max) { round(runif(n, min, max)) }

#' Generate random invertible symmetric pxp matrix
#'
#' @param p 
#' @param M 
#' @param alpha regularization
#'
#' @return
randinvmat <- function(p, M = 3, alpha = 1) {
   mat <- matrix(randi(p^2, -M, M), nrow = p, ncol = p)
   t(mat) %*% mat + diag(p) * alpha
}
```
## Data generation

We will generate the data according to the model.

```{r}
p <- 4      # number of variables
m <- 10     # number of sources
n <- 20     # number of items per source

list_exact <- list()
```

Upper hierarchical level:

```{r}
list_exact$B.exact <- randinvmat(p)      # Between covariance mtx
list_exact$U.exact <- randinvmat(p)      # Inverted Wishart parameter
list_exact$mu.exact <- randi(p, 0, 10)   # Mean of means
list_exact$nw.exact <- 2*(p + 1) + 1     # dof for Inverted Wishart, as small as possible
```

Middle hierarchical level:    

- source means ($\theta_i$) 
- within covariance matrices $W_i$

```{r}
theta.sources.exact <- list()
for (i in seq(m)) {
  theta.sources.exact[[i]] <- mvtnorm::rmvnorm(1, list_exact$mu.exact, list_exact$B.exact)
}
W.sources.exact <- list()
for (i in seq(m)) {
  # W.sources[[i]] <- solve(rWishart(1, nw.exact, solve(U.exact))[,,1])
  W.sources.exact[[i]] <- riwish_Press(list_exact$nw.exact, list_exact$U.exact)
}
```

Lower hierarchical level: raw data

```{r}
df <- list()
for (i in seq(m)) {
  # The data
  df[[i]] <- rmvnorm(n, theta.sources.exact[[i]], W.sources.exact[[i]])
  # The source (last column)
  df[[i]] <- cbind(df[[i]], i)
}

df <- data.frame(do.call(rbind, df))
colnames(df) <- paste(c(paste0('v', seq(1:p)), 'source'))
```

The item column is the last one (`p + 1`).

### Visualization

First two components across sources

```{r}
library(ggplot2)

ggplot(data = df, aes(x = v1, y = v2, col = factor(source))) +
   geom_point(size = 1) +
   scale_colour_discrete('Source')
```

## Prior elicitation

The package supplies the function `make_priors_and_init`.

Suppose that `df` represents a background dataset.  
We can use it to elicit the model hyperparameters:

```{r}

col.variables <- 1:p
col.item <- p + 1

use.priors <- 'ML'
use.init <- 'random'

priors <- make_priors_and_init(df, col.variables, col.item, use.priors, use.init)
names(priors)
```

We obtain:

```{r}
bmatrix(as.matrix(priors$mu), pre = '\\mu =', digits = bmatrix.digits)
bmatrix(priors$B.inv, pre = 'B^{-1} =', digits = bmatrix.digits)
```

Hyperpriors for the within covariance matrices:
$$ W_i \sim IW(n_w, U) $$
$$ n_w = `r priors$nw` $$
```{r}
bmatrix(priors$U, pre = 'U =', digits = bmatrix.digits)
```

Initialisation:    
Within covariance matrices and their inverses:
```{r}
bmatrix(priors$W.inv.1, pre = 'W^{-1}_1 =', digits = bmatrix.digits)
bmatrix(priors$W.inv.2, pre = 'W^{-1}_2 =', digits = bmatrix.digits)
bmatrix(solve(priors$W.inv.1), pre = 'W_1 =', digits = bmatrix.digits)
bmatrix(solve(priors$W.inv.2), pre = 'W_2 =', digits = bmatrix.digits)
```
```

